Estimating Information Transfer in Hard Parameter Sharing via Precise Bias-variance Asymptotics

### 2021-07-14 revision_journal/results.tex
finish sec 3.2

    We first describe a setting for modeling covariate shift.
    Let $\cS$ be the set of positive definite matrices $M^{\top} M$ such that for $i = 1,2,\dots,\floor{\frac p 2}$, $\lambda_{p - i}  = \frac 1 {\lambda_i}$, where $\lambda_1 \ge \lambda_2 \ge \cdots \ge \lambda_p$ are the eigenvalues of $M$.
    Under this setting, we ask: suppose that the sample sizes $n_1$ and $n_2$ are both fixed, which $M$ provides the best ``transfer'' in the sense of HPS estimator's excess risk?
    We observe a dichotomy that depends on whether or not $n_1$ is greater than $n_2$.
    \begin{claim}[Covariate shift vs. sample size ratios]\label{claim_dichotomy}
        Let $g(M) = \frac{\sigma^2}{n_1 + n_2} \bigtr{\frac 1 {\alpha_1 M^{\top} M + \alpha_2}}$.
        Suppose that $n_1$ and $n_2$ are both fixed.
        Within the set of all possible covariate shift matrix $M\in\cS$, the following dichotomy holds:
        \begin{itemize}
	        \item If $n_1 \ge n_2$, then $g(M)$ is minimized in $\cS$ when $\lambda_i = 1$, for any $i = 1,\dots,p$.
	        \item If $n_1 < n_2$, then $g(M)$ is maximized in $\cS$ when $\lambda_i = 1$, for any $i = 1,\dots,p$.
        \end{itemize}
    \end{claim}

\begin{proof}
    We explain why the above dichotomy happens. We can write $g(M)$ for any $M \in \cS$ as
    $$g(M)=\frac{\sigma^2}{ n_1+n_2 }\sum_{i=1}^{p/2}\left( \frac{1}{\lambda_i^{2} a_1 + a_2} + \frac1{\lambda_i^{-2} a_1 + a_2} \right).$$
    When $M=\id_{p\times p}$, by the first equation of \eqref{eq_a12extra}, we have
    $$g(\id_{p\times p})=\frac{\sigma^2}{ n_1+n_2 }\sum_{i=1}^{p/2} \frac{2}{1-\gamma},$$
    where we abbreviate $\gamma:=p/(n_1+n_2)$. Then using $\alpha_1 + \alpha_2 = 1-\gamma$, through a direct calculation we find that
    \begin{align*}
        g(M) - g(\id_{p\times p})%&= \frac{\sigma^2 \gamma}{2(1-\gamma)} (\lambda^2-1)a_1\cdot \bigbrace{  \frac{1}{ -a_1(\lambda^2-1)+(1-\gamma)\lambda^2 } - \frac{1}{a_1(\lambda^2-1) + (1-\gamma)}} \\
        &= \frac{\sigma^2 }{ n_1+n_2-p} \sum_{i=1}^{p/2} \frac{(\lambda_i^2-1)^2 \alpha_1 \left[ \alpha_1 - \alpha_2\right] }{[\alpha_1 + \lambda_i^2 \alpha_2][\lambda_i^2 \alpha_1 + \alpha_2]} .
    \end{align*}
    %in this example is equal to
    %$\frac{p}{2(n_1 + n_2)} f(\lambda)$, where
    %\[ f(\lambda) = {(\lambda^{-2} a_1 + a_2)^{-1} + (\lambda^2 a_1 + a_2)^{-1}}. \]
    %Using the fact that $a_1 + a_2 = 1 - \frac{p}{n_1 + n_2}$, we can verify
    %\begin{align*}
    %	f(\lambda) - f(1) &= \left(\lambda^2 a_1 + \frac{n_1 + n_2 - p}{n_1 + n_2} - a_1\right)^{-1} \\
    %	&+ \left(\lambda^{-2} a_1 + \frac{n_1 + n_2 - p}{n_1 + n_2} - a_1\right)^{-1} \\
    %	&- \frac{2(n_1 + n_2)}{n_1 + n_2 - p} \\
    %	&= \left(2a_1 - \frac{n_1 + n_2-p} {n_1 + n_2 }\right)  g(\lambda, a_1), %\cdot (\lambda^2-1)^2
    %\end{align*}
    %\begin{align*}
    %	f(\lambda) - f(1) &= \left(2a_1 - \frac{n_1 + n_2-p} {n_1 + n_2 }\right)  g(\lambda, a_1), %\cdot (\lambda^2-1)^2
    %\end{align*}
    %where $g(\lambda, a_1) \ge 0$.
    %and can be derived from algebraic calculations (details omitted).
    We claim that $\alpha_1 > \alpha_2$ if and only if $n_1 > n_2$, which then explains the dichotomy.
    In fact, if $\alpha_1 > \alpha_2$, then the first part of equation \eqref{eq_a12extra} gives that $\alpha_1 > (1-\gamma)/2$.
    The second part of equation \eqref{eq_a12extra} gives that
    \begin{align*}
     \frac{n_1}{n_1 + n_2} &> \alpha_1 + \frac{1}{n_1+n_2} \sum_{i=1}^{p/2}\left(\frac{\lambda_i^2}{\lambda_i^2+1}+\frac{\lambda_i^{-2}}{\lambda_i^{-2}+1}\right) = \frac{1-\gamma}{2}+\frac{\gamma}{2}=\frac{1}{2}.
    \end{align*}
    %\begin{align*}
    % \frac{n_1}{n_1 + n_2} &=a_1 + \frac1{n_1 + n_2}\cdot \bigbrace{\sum_{i=1}^p \frac{\lambda_i^2 a_1}{\lambda_i^2 a_1 + a_2}} \\
    %	&> a_1 + \frac{p}{2(n_1+n_2)} \left(\frac{\lambda^2}{\lambda^2+1}+\frac{\lambda^{-2}}{\lambda^{-2}+1}\right) =\frac{1}{2}.
    %\end{align*}
    This is equivalent to $n_1>n_2$. The other direction follows from a similar argument. % Similarly, if $a_1<a_2$, equations  \eqref{eq_a12extra000} and  \eqref{eq_a12extra} give that $a_1 < \frac{n_1 + n_2-p}{2 (n_1 + n_2)}$ and $n_1<n_2$. Thus, we conclude that $f(\lambda) \ge f(1)$ if and only if $n_1 \ge n_2$.
    %\end{example}
\end{proof}

@@

 In this subsection, we consider the setting where $X^{(1)}$ and $X^{(2)}$ are independent and have the same population covariance matrices $\Sigma^{(1)}=\Sigma^{(2)}$. However, the two tasks can have different sample sizes $n_1\ne n_2$. In this case, we can obtain the exact asymptotic limits of the bias term \eqref{Lbias} and the variance term \eqref{Lvar}. We will use these results to illustrate the effects of the bias-variance tradeoff and varying sample sizes on the performance of the $\MTL$ estimator.

\begin{theorem}[Excess risk of HPS under model shift]\label{cor_MTL_loss}
Under Assumption \ref{assm_big1}, suppose that $\Sigma^{(1)}=\Sigma^{(2)}$ and the entries of $Z^{(1)}$ and $Z^{(2)}$ are i.i.d. Gaussian random variables. Then for any small constant $\e>0$ and large constant $C>0$, there exists a high probability event $\Xi$, on which the following estimates hold uniformly in all $a\in \R$:
\be\label{Lvar_samplesize}
L_{\var}(a)= \sigma^2  \left[ \cal L_1(a)+ \OO\left({p^{c}}/{n_1}\right)\right]  ,
\ee
and
\begin{align}
L_{\bias}(a)&= \left[ \cal L_2(a)+  \OO\left(p^{-1/2+c}\right)\right] \cdot  \|\beta^{(1)}-a\beta^{(2)}\|^2 \nonumber\\
&+ \OO \left[ p^{-C}\left( \|\beta^{(1)} \|^2  +  \|\beta^{(2)} \|^2\right)   \right].\label{Lbias_samplesize}
	%-\left[1- \left( 1-\frac{1}{\sqrt{\rho_1}}\right)^4\right] pd^2\cdot \frac{\rho_1^2 (\rho_1+\rho_2)}{(\rho_1 + \rho_2 - 1)^3} +\OO(p^{-c}\sigma^2)  \le
	%   \left|L(\hat{\beta}_2^{\MTL}) - \frac{2d^2 n_1^2 (n_1 + n_2)}{(n_1 + n_2 - p)^3} -\frac{\sigma^2 p}{n_1 + n_2 - p}  \right|
	% \le \varepsilon \cdot \frac{2d^2 n_1^2 (n_1 + n_2)}{(n_1 + n_2 - p)^3} +  \OO(p^{-c/2}).\label{cor_MTL_error}
	%\left[\left( 1+\frac{1}{\sqrt{\rho_1}}\right)^4-1\right] d^2\cdot \frac{\rho_1^2 (\rho_1+\rho_2)}{(\rho_1 + \rho_2 - 1)^3} \\
	%& +C \left[(p^{-c_\varphi}+p^{-c_\infty/2})(\sigma^2 +d^2)+p^{-c_\infty}\kappa^2 + %\frac{d^4+\sigma^2 d^2}{\kappa^2}\right],\nonumber
	 \end{align}
Here we define the functions
\begin{align*}
&\cal L_1(a):= \frac{2p}{  {n_2} (1-\xi_2) + a^2 n_1 (1-\xi_1)+ \sqrt{[ {n_2} (1-\xi_2) +a^2 n_1 (1-\xi_1)]^2 + 4a^2{n_2}{n_1} (\xi_1+\xi_2 -\xi_1\xi_2)}}, \\
&\cal L_2(a): = \frac1{a^2}\cdot \frac{1- 2\frac{\cal L_1(a)}{\xi_2[1 + \cal L_1(a)]} + \kappa(a)}{1- \xi_2 \kappa(a)},
\end{align*}
where we abbreviate $\xi_1:=p/n_1$, $\xi_2:=p/n_2$ and
\begin{align*}
\kappa(a):= \frac{\cal L_1(a)^2}{\xi_2^2[1+\cal L_1(a)]^2}  \left[1 - \frac{a^4 \cal L_1(a)^2 }{\xi_1[1+a^2\cal L_1(a) ]^2}\right]^{-1}.
%f_2(a):= \frac{a^4 \cal L_1(a)^2}{\xi_1^2 } \left[1 - \frac{a^4 \cal L_1(a)^2 }{\xi_1[1+a^2\cal L_1(a) ]^2}\right]^{-1},\\
%f_3(a):= \frac1{a^2}\frac{\xi_1}{\xi_2 +\cal L_1(a)}
 \end{align*}
%	In the setting of Example \ref{ex_sample_ratio}, assume that
%	%a) the sample sizes $n_1$ and $n_2$ are greater than $(1 + \tau) p$, b) $\Sigma_1=\Sigma_2=\id_p$, and c) %there exists a small constant $c_0>0$ such that
%	(i) both tasks sample sizes are at least $3p$;
%	(ii) noise variance is smaller than the shared signal variance: $\sigma^2 \lesssim  \kappa^2$;
%%	\be\label{choiceofpara0}
%%	p^{-1/2+c_0}\sigma^2 + p^{c_0}d^2\le \kappa^2\le p^{1-c_0} (\sigma^2 +d^2)  .  	\ee
%	%\be\label{choiceofpara0}
%%	(ii) the task-specific variance of $\beta_i$ is much smaller than the signal strength {\color{red}$d^2 = \oo( {\kappa^2})$}; \HZ{what does $\ll$ mean exactly?}
%%	(iii) the sample sizes $n_1$ and $n_2$ are greater than $(1 + \tau) p$.
%	(iii) task-specific variance is much smaller than the shared signal variance: $d^2 \le p^{-\e}{\kappa^2}$ for a small constant $c>0$.
%	Let $\varepsilon = (1 + \sqrt{p/n_1})^ 4 - 1$, which decreases as $n_1$ increases.
%	Let $\hat{A},\hat{B}$ be the global minimizer of $f(A, B)$.
%	With high probability over the randomness of the input,
%	the prediction loss of $\hat{\beta}_2^{\MTL} = \hat{B} \hat{A}_2$ for task two satisfies that
%	\begin{align}
%	%-\left[1- \left( 1-\frac{1}{\sqrt{\rho_1}}\right)^4\right] pd^2\cdot \frac{\rho_1^2 (\rho_1+\rho_2)}{(\rho_1 + \rho_2 - 1)^3} +\OO(p^{-c}\sigma^2)  \le
%	   \left|L(\hat{\beta}_2^{\MTL}) - \frac{2d^2 n_1^2 (n_1 + n_2)}{(n_1 + n_2 - p)^3} -\frac{\sigma^2 p}{n_1 + n_2 - p}  \right|
%	\le \varepsilon \cdot \frac{2d^2 n_1^2 (n_1 + n_2)}{(n_1 + n_2 - p)^3} +  \OO(p^{-c/2}).\label{cor_MTL_error}
%	%\left[\left( 1+\frac{1}{\sqrt{\rho_1}}\right)^4-1\right] d^2\cdot \frac{\rho_1^2 (\rho_1+\rho_2)}{(\rho_1 + \rho_2 - 1)^3} \\
%	%& +C \left[(p^{-c_\varphi}+p^{-c_\infty/2})(\sigma^2 +d^2)+p^{-c_\infty}\kappa^2 + %\frac{d^4+\sigma^2 d^2}{\kappa^2}\right],\nonumber
%	 \end{align}
%	 with high probability for any fixed $c\in(0, \min(\frac{1}{4}, \delta,\frac{\varphi-4}{2\varphi}))$.
%	 {\color{red}[FY: the error also contains $p^{-1/2+2c}\kappa^2 +  p^{-1/4+c} (\sigma^2 +d^2) $, both of which cannot be omitted, because (i) there is no assumption on the upper bound of $\kappa^2$, and (ii) we do not necessarily have $c_\varphi<1/4$. We can decide how to present the result concisely (for instance we can impose an upper bound on $\kappa^2$ and that $c_\varphi<1/4$), but it needs to be correct.]}
	 \end{theorem}

@@

We believe the convergence rates $p^c/n_1$ and $p^{-1/2+c}$ in \eqref{Lvar_samplesize} and \eqref{Lbias_samplesize} are both sharp up to the $p^c$ factor. The proof of Theorem \ref{cor_MTL_loss} will be given in Appendix \ref{app_iso_cov} in the supplement \cite{MTL_suppl}. For the variance estimate in \eqref{Lvar_samplesize}, it is not necessary to assume the Gaussian distributions of the $Z^{(1)}$ and $Z^{(2)}$ entries. In fact, \eqref{Lvar_samplesize} is a special case of Theorem \ref{thm_main_RMT} below for the more general case with possibly different $\Sigma^{(1)}$ and $\Sigma^{(2)}$. On the other hand, the Gaussian assumption is needed in our current proof of the bias limit \eqref{Lbias_samplesize}. In the setting of Theorem \ref{cor_MTL_loss}, we can write
\begin{align*}
L_{\bias} (a)&=\bv_a^\top (Z^{(1)})^\top Z^{(1)} \frac1{ \left[a^2(Z^{(1)})^\top Z^{(1)}+ (Z^{(2)})^\top Z^{(2)}  \right]^{2}}(Z^{(1)})^\top Z^{(1)} \bv_a,
\end{align*}
where we abbreviate $\bv(a):=(\Sigma^{(1)})^{1/2}\left(a\beta^{(1)}- a^2\beta^{(2)}\right)$. By the rotational invariance of $(Z^{(1)})^\top Z^{(1)}$ and $(Z^{(2)})^\top Z^{(2)}$, we have that
\begin{align}\label{Lbias_idea}
L_{\bias} (a)&\approx \|\bv_a\|^2 \frac1p\bigtr{ [(Z^{(1)})^\top Z^{(1)}]^2 \frac1{ \left[a^2(Z^{(1)})^\top Z^{(1)}+ (Z^{(2)})^\top Z^{(2)}\right]^{2}}  },
\end{align}
up to a small error. Notice that we can write \eqref{Lbias_idea} into a simpler form
\begin{align}\nonumber %\label{Lbias_idea2}
L_{\bias} (a)&\approx \|\bv_a\|^2 \left. \frac{\dd }{\dd x}\right|_{x=0}\frac1p\bigtr{  \frac1{ a^2(Z^{(1)})^\top Z^{(1)}+ x[(Z^{(1)})^\top Z^{(1)}]^2+(Z^{(2)})^\top Z^{(2)} }  }.
\end{align}
It is well-known that the empirical spectral distributions (ESD) of $(Z^{(i)})^\top Z^{(i)}$, $i=1,2$, satisfy the famous Marchenko-Pastur (MP) law asymptotically \cite{MP}. From the MP law of $(Z^{(1)})^\top Z^{(1)}$, we can also derive the asymptotic ESD of $a^2(Z^{(1)})^\top Z^{(1)}+ x[(Z^{(1)})^\top Z^{(1)}]^2$ for any fixed $a\in \R$ and $x>0$. Due to the rotational invariance of multivariate Gaussian distributions,
%$a^2(Z^{(1)})^\top Z^{(1)}+ x[(Z^{(1)})^\top Z^{(1)}]^2$ and $(Z^{(2)})^\top Z^{(2)}$ are asymptotically freely independent from each other \cite{nica2006lectures}. Hence
the asymptotic ESD of $a^2(Z^{(1)})^\top Z^{(1)}+ x[(Z^{(1)})^\top Z^{(1)}]^2+(Z^{(2)})^\top Z^{(2)}$ is given by the free additive convolution (or free addition) of the asymptotic ESD of $a^2(Z^{(1)})^\top Z^{(1)}+ x[(Z^{(1)})^\top Z^{(1)}]^2$ and the MP law of $(Z^{(2)})^\top Z^{(2)}$\cite{nica2006lectures}. In particular, a sharp convergence estimate has been proved in \cite{BES_free1,BES_free2} for the free addition of two probability measures. We use that result to obtain a convergence estimate on
$$\frac1p\bigtr{  \frac1{ a^2(Z^{(1)})^\top Z^{(1)}+ x[(Z^{(1)})^\top Z^{(1)}]^2+(Z^{(2)})^\top Z^{(2)} }  }.$$
Taking derivative with respect to $x$ at $x=0$ gives the exact asymptotic limit of $L_{\bias}(a)$. We refer the reader to Appendix \ref{app_iso_cov} of the supplement \cite{MTL_suppl} for more details.

We believe that the above argument can be extended to the case without the Gaussian assumption. For example, instead of using the results in \cite{BES_free1,BES_free2} on free addition, we can use the sharp local laws on polynomials of random matrices in \cite{EKN_poly}. However, to apply the result in \cite{EKN_poly}, we need to check some difficult technical regularity conditions for our setting, which is not the focus of this work. Hence we do not pursue this direction in this paper.

>>>>

    We first describe a setting for modeling covariate shift.
    Let $\cS$ be the set of positive definite matrices $M^{\top} M$ such that for $i = 1,2,\dots,\floor{\frac p 2}$, $\lambda_{p - i}  = \frac 1 {\lambda_i}$, where $\lambda_1 \ge \lambda_2 \ge \cdots \ge \lambda_p$ are the eigenvalues of $M$.
    Under this setting, we ask: suppose that the sample sizes $n_1$ and $n_2$ are both fixed, which $M$ provides the best ``transfer'' in the sense of HPS estimator's excess risk?
    We observe a dichotomy that depends on whether or not $n_1$ is greater than $n_2$.
    \begin{claim}[Sample sizes vs. covariate shift]\label{claim_dichotomy}
        Let $g(M) = \frac{\sigma^2}{n_1 + n_2} \bigtr{\frac 1 {\alpha_1 M^{\top} M + \alpha_2}}$.
        Suppose that $n_1$ and $n_2$ are both fixed.
        Within the set of all possible covariate shift matrix $M\in\cS$, the following dichotomy holds:
        \begin{enumerate}
	        \item[i)] If $n_1 \ge n_2$, then $g(M)$ is minimized in $\cS$ when $\lambda_i = 1$, for any $i = 1,\dots,p$.
	        \item[ii)] If $n_1 < n_2$, then $g(M)$ is maximized in $\cS$ when $\lambda_i = 1$, for any $i = 1,\dots,p$.
        \end{itemize}
    \end{claim}

\begin{proof}
    For any $M \in \cS$, we can write $g(M)$ as
    $$g(M)=\frac{\sigma^2}{ n_1+n_2 }\sum_{i=1}^{p/2}\left( \frac{1}{\lambda_i^{2} \alpha_1 + \alpha_2} + \frac1{\lambda_i^{-2} \alpha_1 + \alpha_2} \right).$$
    When $M=\id_{p\times p}$, by the first equation of \eqref{eq_a12extra}, we have
    $$g(\id_{p\times p})=\frac{\sigma^2}{ n_1+n_2 }\sum_{i=1}^{p/2} \frac{2}{1-\gamma},$$
    where we abbreviate $\gamma:=p/(n_1+n_2)$.
    Using $\alpha_1 + \alpha_2 = 1-\gamma$, we further have that
    \begin{align*}
        g(M) - g(\id_{p\times p})%&= \frac{\sigma^2 \gamma}{2(1-\gamma)} (\lambda^2-1)a_1\cdot \bigbrace{  \frac{1}{ -a_1(\lambda^2-1)+(1-\gamma)\lambda^2 } - \frac{1}{a_1(\lambda^2-1) + (1-\gamma)}} \\
        &= \frac{\sigma^2 }{ n_1+n_2-p} \sum_{i=1}^{p/2} \frac{(\lambda_i^2-1)^2 \alpha_1 \left[ \alpha_1 - \alpha_2\right] }{[\alpha_1 + \lambda_i^2 \alpha_2][\lambda_i^2 \alpha_1 + \alpha_2]} .
    \end{align*}
    %in this example is equal to
    %$\frac{p}{2(n_1 + n_2)} f(\lambda)$, where
    %\[ f(\lambda) = {(\lambda^{-2} a_1 + a_2)^{-1} + (\lambda^2 a_1 + a_2)^{-1}}. \]
    %Using the fact that $a_1 + a_2 = 1 - \frac{p}{n_1 + n_2}$, we can verify
    %\begin{align*}
    %	f(\lambda) - f(1) &= \left(\lambda^2 a_1 + \frac{n_1 + n_2 - p}{n_1 + n_2} - a_1\right)^{-1} \\
    %	&+ \left(\lambda^{-2} a_1 + \frac{n_1 + n_2 - p}{n_1 + n_2} - a_1\right)^{-1} \\
    %	&- \frac{2(n_1 + n_2)}{n_1 + n_2 - p} \\
    %	&= \left(2a_1 - \frac{n_1 + n_2-p} {n_1 + n_2 }\right)  g(\lambda, a_1), %\cdot (\lambda^2-1)^2
    %\end{align*}
    %\begin{align*}
    %	f(\lambda) - f(1) &= \left(2a_1 - \frac{n_1 + n_2-p} {n_1 + n_2 }\right)  g(\lambda, a_1), %\cdot (\lambda^2-1)^2
    %\end{align*}
    %where $g(\lambda, a_1) \ge 0$.
    %and can be derived from algebraic calculations (details omitted).
    We claim that $\alpha_1 > \alpha_2$ if and only if $n_1 > n_2$, which proves claim i).
    In fact, if $\alpha_1 > \alpha_2$, then the first part of equation \eqref{eq_a12extra} gives that $\alpha_1 > (1-\gamma)/2$.
    The second part of equation \eqref{eq_a12extra} gives that
    \begin{align*}
     \frac{n_1}{n_1 + n_2} &> \alpha_1 + \frac{1}{n_1+n_2} \sum_{i=1}^{p/2}\left(\frac{\lambda_i^2}{\lambda_i^2+1}+\frac{\lambda_i^{-2}}{\lambda_i^{-2}+1}\right) = \frac{1-\gamma}{2}+\frac{\gamma}{2}=\frac{1}{2},
    \end{align*}
    %\begin{align*}
    % \frac{n_1}{n_1 + n_2} &=a_1 + \frac1{n_1 + n_2}\cdot \bigbrace{\sum_{i=1}^p \frac{\lambda_i^2 a_1}{\lambda_i^2 a_1 + a_2}} \\
    %	&> a_1 + \frac{p}{2(n_1+n_2)} \left(\frac{\lambda^2}{\lambda^2+1}+\frac{\lambda^{-2}}{\lambda^{-2}+1}\right) =\frac{1}{2}.
    %\end{align*}
    which is equivalent to $n_1>n_2$. The proof of claim ii) follows from a similar argument. % Similarly, if $a_1<a_2$, equations  \eqref{eq_a12extra000} and  \eqref{eq_a12extra} give that $a_1 < \frac{n_1 + n_2-p}{2 (n_1 + n_2)}$ and $n_1<n_2$. Thus, we conclude that $f(\lambda) \ge f(1)$ if and only if $n_1 \ge n_2$.
    %\end{example}
\end{proof}

@@

Next, we consider a model shift setting where task one and two have the same linear model but different population covariance matrices.
We show the exact asymptotic limits of the bias and variance of HPS.

\begin{theorem}[Excess risk of HPS under model shift]\label{cor_MTL_loss}
Under Assumption \ref{assm_big1}, suppose that $\Sigma^{(1)}=\Sigma^{(2)}$ and the entries of $Z^{(1)}$ and $Z^{(2)}$ are i.i.d. Gaussian random variables.
Denote by $\xi_1:=p/n_1$ and $\xi_2:=p/n_2$.
Then, for any small constant $\e>0$ and large constant $C>0$, there exists a high probability event $\Xi$, on which the following estimates hold uniformly in all $a\in \R$:
\begin{align}
L_{\var}(a) &= \sigma^2  \cal L_1(a)+ \OO\Big(\frac {\sigma^2 p^{c}} {n_1}\Big), \label{Lvar_samplesize} \\
L_{\bias}(a)&= \bignorm{\beta^{(1)}-a\beta^{(2)}}^2 \cal L_2(a) +  \OO\Big(p^{-\frac 1 2 + c}  \norm{\beta^{(1)}-a\beta^{(2)}}^2
+ p^{-C}\big( \|\beta^{(1)} \|^2  +  \|\beta^{(2)} \|^2\big)   \Big),\label{Lbias_samplesize}
\end{align}
where $\cL_1(a)$ and $\cL_2(a)$ are defined as follows
\begin{align*}
\cal L_1(a) &:= {2p}\cdot\left((n_2 - p) + a^2 (n_1 - p) + \sqrt{\big( (n_2 - p) + a^2 (n_1 - p)\big)^2 + 4a^2(n_1 p + n_2 p - n_1 n_2)}\right)^{-1}, \\
\cal L_2(a) &:= \frac1{a^2}\cdot \frac{1- 2\frac{\cal L_1(a)}{\xi_2(1 + \cal L_1(a))} + \kappa(a)}{1- \xi_2 \kappa(a)}, \text{ in which }
\kappa(a) := \frac{\cal L_1(a)^2}{\xi_2^2(1+\cal L_1(a))^2}  \left(1 - \frac{a^4 \cal L_1(a)^2 }{\xi_1(1+a^2\cal L_1(a) )^2}\right)^{-1}.
%f_2(a):= \frac{a^4 \cal L_1(a)^2}{\xi_1^2 } \left[1 - \frac{a^4 \cal L_1(a)^2 }{\xi_1[1+a^2\cal L_1(a) ]^2}\right]^{-1},\\
%f_3(a):= \frac1{a^2}\frac{\xi_1}{\xi_2 +\cal L_1(a)}
 \end{align*}
%	In the setting of Example \ref{ex_sample_ratio}, assume that
%	%a) the sample sizes $n_1$ and $n_2$ are greater than $(1 + \tau) p$, b) $\Sigma_1=\Sigma_2=\id_p$, and c) %there exists a small constant $c_0>0$ such that
%	(i) both tasks sample sizes are at least $3p$;
%	(ii) noise variance is smaller than the shared signal variance: $\sigma^2 \lesssim  \kappa^2$;
%%	\be\label{choiceofpara0}
%%	p^{-1/2+c_0}\sigma^2 + p^{c_0}d^2\le \kappa^2\le p^{1-c_0} (\sigma^2 +d^2)  .  	\ee
%	%\be\label{choiceofpara0}
%%	(ii) the task-specific variance of $\beta_i$ is much smaller than the signal strength {\color{red}$d^2 = \oo( {\kappa^2})$}; \HZ{what does $\ll$ mean exactly?}
%%	(iii) the sample sizes $n_1$ and $n_2$ are greater than $(1 + \tau) p$.
%	(iii) task-specific variance is much smaller than the shared signal variance: $d^2 \le p^{-\e}{\kappa^2}$ for a small constant $c>0$.
%	Let $\varepsilon = (1 + \sqrt{p/n_1})^ 4 - 1$, which decreases as $n_1$ increases.
%	Let $\hat{A},\hat{B}$ be the global minimizer of $f(A, B)$.
%	With high probability over the randomness of the input,
%	the prediction loss of $\hat{\beta}_2^{\MTL} = \hat{B} \hat{A}_2$ for task two satisfies that
%	\begin{align}
%	%-\left[1- \left( 1-\frac{1}{\sqrt{\rho_1}}\right)^4\right] pd^2\cdot \frac{\rho_1^2 (\rho_1+\rho_2)}{(\rho_1 + \rho_2 - 1)^3} +\OO(p^{-c}\sigma^2)  \le
%	   \left|L(\hat{\beta}_2^{\MTL}) - \frac{2d^2 n_1^2 (n_1 + n_2)}{(n_1 + n_2 - p)^3} -\frac{\sigma^2 p}{n_1 + n_2 - p}  \right|
%	\le \varepsilon \cdot \frac{2d^2 n_1^2 (n_1 + n_2)}{(n_1 + n_2 - p)^3} +  \OO(p^{-c/2}).\label{cor_MTL_error}
%	%\left[\left( 1+\frac{1}{\sqrt{\rho_1}}\right)^4-1\right] d^2\cdot \frac{\rho_1^2 (\rho_1+\rho_2)}{(\rho_1 + \rho_2 - 1)^3} \\
%	%& +C \left[(p^{-c_\varphi}+p^{-c_\infty/2})(\sigma^2 +d^2)+p^{-c_\infty}\kappa^2 + %\frac{d^4+\sigma^2 d^2}{\kappa^2}\right],\nonumber
%	 \end{align}
%	 with high probability for any fixed $c\in(0, \min(\frac{1}{4}, \delta,\frac{\varphi-4}{2\varphi}))$.
%	 {\color{red}[FY: the error also contains $p^{-1/2+2c}\kappa^2 +  p^{-1/4+c} (\sigma^2 +d^2) $, both of which cannot be omitted, because (i) there is no assumption on the upper bound of $\kappa^2$, and (ii) we do not necessarily have $c_\varphi<1/4$. We can decide how to present the result concisely (for instance we can impose an upper bound on $\kappa^2$ and that $c_\varphi<1/4$), but it needs to be correct.]}
	 \end{theorem}

@@

Combining equation \eqref{Lvar_samplesize} and \eqref{Lbias_samplesize}, we thus obtain an estimate for the excess risk of HPS.
We provide some intuition behind both estimates.
%For the variance estimate in \eqref{Lvar_samplesize}, it is not necessary to assume the Gaussian distributions of the $Z^{(1)}$ and $Z^{(2)}$ entries.
First, the variance estimate is a special case of Theorem \ref{thm_main_RMT} where $\Sigma^{(1)} = \Sigma^{(2)}$ but $n_1 \neq n_2$.
In fact, the result can be obtained by solving $\alpha_1, \alpha_2$ in equation \eqref{eq_a12extra} using $\lambda_i = 1$ for all $i = 1,\dots, p$.
%On the other hand, the Gaussian assumption is needed in our current proof of the bias limit \eqref{Lbias_samplesize}.
Second, the bias estimate is obtained based on a sharp convergence estimate proved in \citet{BES_free1,BES_free2} for the free addition of two probability measures.
To illustrate, denote by $\bv(a):=(\Sigma^{(1)})^{1/2}\left(a\beta^{(1)}- a^2\beta^{(2)}\right)$.
We can write the bias equation as
\begin{align*}
    L_{\bias} (a)&=\bv_a^\top (Z^{(1)})^\top Z^{(1)} { \left(a^2(Z^{(1)})^\top Z^{(1)}+ (Z^{(2)})^\top Z^{(2)}  \right)^{-2}}(Z^{(1)})^\top Z^{(1)} \bv_a,
\end{align*}
By the rotational invariance of $(Z^{(1)})^\top Z^{(1)}$ and $(Z^{(2)})^\top Z^{(2)}$, we have that
\begin{align}\label{Lbias_idea}
    L_{\bias} (a)&\approx \|\bv_a\|^2 \cdot \frac1p\bigtr{ \big((Z^{(1)})^\top Z^{(1)}\big)^2 { \left(a^2(Z^{(1)})^\top Z^{(1)}+ (Z^{(2)})^\top Z^{(2)}\right)^{-2}}}
\end{align}
up to a small error.

Our key idea is to write equation \eqref{Lbias_idea} as the derivative of a certain polynomial with respect to $x$ at $x = 0$:
\begin{align}\nonumber %\label{Lbias_idea2}
L_{\bias} (a)&\approx \|\bv_a\|^2 \cdot\left. \frac{\dd }{\dd x}\right|_{x=0}\frac1p\bigtr{  \Big( a^2(Z^{(1)})^\top Z^{(1)} + x \big((Z^{(1)})^\top Z^{(1)}\big)^2 + (Z^{(2)})^\top Z^{(2)} \Big)^{-1}  }.
\end{align}
It is well-known that the empirical spectral distributions (ESD) of $(Z^{(i)})^\top Z^{(i)}$, $i=1,2$, satisfy the famous Marchenko-Pastur (MP) law asymptotically \cite{MP}. From the MP law of $(Z^{(1)})^\top Z^{(1)}$, we can also derive the asymptotic ESD of $a^2(Z^{(1)})^\top Z^{(1)}+ x[(Z^{(1)})^\top Z^{(1)}]^2$ for any fixed $a\in \R$ and $x>0$. Due to the rotational invariance of multivariate Gaussian distributions,
%$a^2(Z^{(1)})^\top Z^{(1)}+ x[(Z^{(1)})^\top Z^{(1)}]^2$ and $(Z^{(2)})^\top Z^{(2)}$ are asymptotically freely independent from each other \cite{nica2006lectures}. Hence
the asymptotic ESD of $a^2(Z^{(1)})^\top Z^{(1)}+ x[(Z^{(1)})^\top Z^{(1)}]^2+(Z^{(2)})^\top Z^{(2)}$ is given by the free additive convolution (or free addition) of the asymptotic ESD of $a^2(Z^{(1)})^\top Z^{(1)}+ x[(Z^{(1)})^\top Z^{(1)}]^2$ and the MP law of $(Z^{(2)})^\top Z^{(2)}$\cite{nica2006lectures}.
%In particular, a sharp convergence estimate has been proved in \cite{BES_free1,BES_free2} for the free addition of two probability measures.
Finally, We use the result of \citet{BES_free1,BES_free2} to obtain a sharp estimate of
$$\frac1p\bigtr{  \Big( a^2(Z^{(1)})^\top Z^{(1)}+ x \big((Z^{(1)})^\top Z^{(1)}\big)^2+(Z^{(2)})^\top Z^{(2)} \Big)^{-1}},$$
which implies a sharp estimate of $L_{\bias}(a)$.
%Taking the derivative with respect to $x$ at $x=0$ gives the exact asymptotic limit of $L_{\bias}(a)$.
The complete proof of Theorem \ref{cor_MTL_loss} can be found in Section \ref{app_iso_cov}.

\medskip
\noindent\textit{Remarks.}
First, we believe the convergence rates $p^c/n_1$ and $p^{-1/2+c}$ in \eqref{Lvar_samplesize} and \eqref{Lbias_samplesize} are both sharp up to the $p^c$ factor.
\todo{the rate in \eqref{Lvar_samplesize} seems different from Thm 3.1?}
Second, we believe that the above argument can be extended to the case without the Gaussian assumption. For example, instead of using the results in \cite{BES_free1,BES_free2} on free addition, we can use the sharp local laws on polynomials of random matrices in \cite{EKN_poly}. However, applying the result of \citet{EKN_poly} requires checking certain technical regularity conditions for our setting. We leave this as an open question for future work.
